# Your first model {#sec-first-model}

```{r}
#| include: false
source("R/_setup.R")
```

::: {.callout-note title="What you'll be able to do"}
Run a full eco-evolutionary projection with one call to `MPM()`, find your way around its output,
and turn that output into the three plots you will use for the rest of the workshop.
:::

## One function, one projection

The whole model is driven by a single function, `MPM()`. The model is parameterised using the
species life cycle (the vital rates), the genetic architecture of the trait (phenotypic variance and
heritability), and the relationship between the trait and demographic performance through selection.
Let's project **species 1** (a fast, grass-like life history) with selection acting on fertility —
the package defaults.

```{r}
#| label: first-run
out <- MPM(
  h2   = 0.2,            # heritability of the selected trait
  Vp   = 1,             # phenotypic variance
  Beta = 0.15,          # strength of selection (on the link scale)
  Fe   = species$F[1],      # fertility
  SA   = species$SA[1],     # adult survival
  SJ   = species$SJ[1],     # juvenile survival
  S0   = species$SJ[1],     # newborn survival (here equal to SJ)
  Y    = species$gamma[1],  # maturation rate
  selectedrate = "Fe"   # the vital rate selection acts on
)
```

That's it — a deterministic eco-evolutionary projection over 100 years. In the pre-breeding life
cycle used here, fertility is defined as $F = f\,S_0$, where $S_0$ is survival from birth to the next
census. For simplicity we assume $S_0 = S_J$, although the model can accommodate different values for
newborn and juvenile survival (see the appendices of @vandewalle2025 for details).

## What came back

```{r}
#| label: structure
str(out, max.level = 1)
```

Three pieces matter today:

- **`properties_per_year`** — a tidy data frame with one row per year. It holds population size `N`,
  growth rate `lambda`, the realised variances `Va`/`Vp`, and the mean breeding value of adults and
  juveniles (`M_EBVSA`, `M_EBVSJ`). This is your demographic *and* evolutionary timeline in one table.
- **`pheno_distribution`** — the full trait distribution: the number of individuals in each phenotype
  class, every year. This is what lets us *watch* the trait evolve.
- The scalars **`T_GEN`** (generation time), **`LAMBDA0`** (initial growth rate) and **`V`** (the mean
  change in breeding value per year — a realised adaptation rate).

```{r}
#| label: head-properties
head(out$properties_per_year[, c("year", "N", "lambda", "Va", "M_EBVSA", "M_EBVSJ")])
```

## Plot 1 — the population grows

```{r}
#| label: fig-popsize
#| fig-cap: "Population size over 100 years. With directional selection and no density dependence, the population grows without bound — that is expected, not a bug."
ggplot(out$properties_per_year, aes(year, N)) +
  geom_line(linewidth = 0.9, colour = species_pal["1"]) +
  labs(x = "Year", y = "Population size", title = "Species 1 — population trajectory")
```

::: {.callout-warning title="Heads-up — why the population grows without bound"}
The model does not include density dependence, so positive selection can increase population growth
above replacement and lead to exponential growth. This is intentional: the example is designed to
illustrate evolutionary and demographic responses to selection without the additional effects of
population regulation. If you want a constant-size baseline, set `enforce_initial_lambda1 = TRUE`
(it tunes fertility so the starting $\lambda = 1$).
:::

## Plot 2 — the trait evolves

Here is the evolution half of the story: the mean **breeding value** increases over time as selection
favours individuals with higher trait values.

::: {.callout-important title="Your turn — predict first"}
Before running the code below: with positive selection on fertility, do you expect the mean breeding
value to go **up**, **down**, or **stay flat**? Should adults and juveniles differ?
:::

```{r}
#| label: fig-breeding-value
#| fig-cap: "Mean breeding value of adults and juveniles. Selection shifts the genetic mean upward; the juvenile mean responds first because juveniles are the most recently recruited, more strongly selected cohort."
out$properties_per_year |>
  select(year, adult = M_EBVSA, juvenile = M_EBVSJ) |>
  pivot_longer(-year, names_to = "age_class", values_to = "mean_bv") |>
  ggplot(aes(year, mean_bv, colour = age_class)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = age_pal, name = NULL) +
  labs(x = "Year", y = "Mean breeding value", title = "Genetic change through time")
```

## Plot 3 — the whole distribution sliding

The mean alone hides part of the story: the entire trait distribution shifts over time. Plotting a
few yearly snapshots makes the movement visible. We **normalise** each year's distribution so it sums
to 1, which isolates the shift in trait values from the change in total population size.

```{r}
#| label: fig-pheno-distribution
#| fig-cap: "Phenotype distribution every 20 years, normalised so each year's curve sums to 1. This shows the shift in trait values independently of changes in total population size."
out$pheno_distribution |>
  group_by(year) |>
  mutate(density_norm = density / sum(density)) |>
  ungroup() |>
  filter(year %in% seq(1, 101, by = 20)) |>
  ggplot(aes(phenotypic_midvalue, density_norm, colour = year, group = year)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_c(option = "mako", direction = -1, name = "Year") +
  labs(x = "Phenotype value", y = "Normalised density",
       title = "The phenotypic distribution shifts over time")
```

::: {.callout-tip title="Key idea"}
`properties_per_year` is your *summary* timeline; `pheno_distribution` is the *full* picture. Almost
every figure in this workshop is one of these two objects, reshaped and plotted.
:::

## Your turn — drive it yourself

::: {.callout-important title="Exercise"}
1. Re-run the projection for the **albatross** (species 5) instead of the grass. Which `species$`
   columns change?
2. Then set `Beta = 0` and re-run. Predict what the breeding-value plot will look like before you draw it.
3. The model uses a pre-breeding census by default (`census = "prebreeding"`). Re-run species 1 with
   `census = "postbreeding"` and compare. Because fertility is built differently in the two life
   cycles (recall $F = f\,S_0$ above), the same fecundity translates into a different fertility — see
   how this changes the projection.
:::

::: {.callout-caution title="Solution" collapse="true"}
```{r}
#| label: solution-ch2
#| fig-height: 3.2
alba <- MPM(
  h2 = 0.2, Vp = 1, Beta = 0.15,
  Fe = species$F[5], SA = species$SA[5], SJ = species$SJ[5],
  S0 = species$SJ[5], Y = species$gamma[5], selectedrate = "Fe"
)

no_sel <- MPM(
  h2 = 0.2, Vp = 1, Beta = 0,            # selection switched off
  Fe = species$F[5], SA = species$SA[5], SJ = species$SJ[5],
  S0 = species$SJ[5], Y = species$gamma[5], selectedrate = "Fe"
)

no_sel$properties_per_year |>
  ggplot(aes(year, M_EBVSA)) +
  geom_line(linewidth = 0.9, colour = "grey40") +
  ylim(-4, 4) +
  labs(x = "Year", y = "Mean breeding value (adults)",
       title = "No selection \u2192 no genetic change")
```

With `Beta = 0` the mean breeding value stays flat: no selection, no adaptation. This is the
sanity check every eco-evolutionary model should pass.

For the census comparison, switch the one argument and re-run:

```{r}
#| label: solution-ch2-census
#| eval: false
post <- MPM(
  h2 = 0.2, Vp = 1, Beta = 0.15,
  Fe = species$F[1], SA = species$SA[1], SJ = species$SJ[1],
  S0 = species$SJ[1], Y = species$gamma[1],
  selectedrate = "Fe", census = "postbreeding"
)
```

Under a post-breeding census the timing of reproduction relative to the census differs, so fertility
is constructed from fecundity and survival in a different way. Comparing the two runs is a concrete
way to see how a life-cycle assumption, not just a parameter value, shapes the results.
:::
