# Selection scenarios {#sec-selection-scenarios}

```{r}
#| include: false
source("R/_setup.R")
```

::: {.callout-note title="What you'll be able to do"}
Put selection on each of the four vital rates in turn, run all the scenarios with a small helper,
and read off *which* vital rate produces the fastest adaptation for a given species.
:::

Selection does not have to act on fertility. The `selectedrate` argument lets you choose which vital
rate the trait influences: `"Fe"` (fertility), `"SJ"` (juvenile survival), `"SA"` (adult survival) or
`"Y"` (maturation). Everything else stays the same — so any difference we see is *about the life
cycle position of the trait*, not about the genetics.

We'll work with **species 3** (a passerine-like, intermediate life history) so all four scenarios are
well behaved.

## A helper to run one scenario

Rather than copy-paste the call four times, wrap it once and vary only `selectedrate`:

```{r}
#| label: scenario-helper
run_scenario <- function(rate, sp = 3) {
  MPM(
    h2 = 0.2, Vp = 1, Beta = 0.15,
    Fe = species$F[sp], SA = species$SA[sp], SJ = species$SJ[sp],
    S0 = species$SJ[sp], Y = species$gamma[sp],
    selectedrate = rate
  )
}

rates <- c("Fe", "SJ", "SA", "Y")
runs  <- set_names(rates) |> map(run_scenario)   # one MPM run per vital rate
```

`runs` is now a named list of four projections. We bind their yearly timelines into one tidy frame:

```{r}
#| label: bind-runs
prop_all <- imap_dfr(runs, \(out, rate)
  out$properties_per_year |> mutate(selected = rate)
)
```

## Which rate moves the trait fastest?

::: {.callout-important title="Your turn — predict first"}
Same species, same heritability, same selection strength — only the *target* vital rate differs.
Which of fertility, juvenile survival, adult survival or maturation do you think will drive the
**fastest** genetic change per year? Commit to a ranking before you look.
:::

```{r}
#| label: fig-bv-by-rate
#| fig-cap: "Mean breeding value through time for species 3, under selection on each of the four vital rates. The steeper the climb, the faster the adaptation."
ggplot(prop_all, aes(year, M_EBVS, colour = selected)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = vital_pal, labels = vital_lab, name = "Selection on") +
  labs(x = "Year", y = "Mean breeding value",
       title = "Same population, four places for selection to act")
```

The slopes differ markedly. To compare them as numbers, each run already reports the realised
adaptation rate per year in `out$V`:

```{r}
#| label: tbl-adapt-rate
adapt_rate <- tibble(
  selected     = rates,
  rate_label   = vital_lab[rates],
  per_year     = map_dbl(runs, "V"),
  gen_time     = map_dbl(runs, "T_GEN")
) |>
  mutate(per_generation = per_year * gen_time) |>
  arrange(desc(per_year))

adapt_rate
```

```{r}
#| label: fig-adapt-rate
#| fig-cap: "Realised adaptation rate per year by selected vital rate (species 3)."
adapt_rate |>
  ggplot(aes(reorder(rate_label, per_year), per_year, fill = selected)) +
  geom_col(width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = vital_pal, guide = "none") +
  labs(x = NULL, y = "Adaptation rate per year (\u0394 breeding value)",
       title = "Selection on juvenile survival adapts fastest here")
```

::: {.callout-tip title="Key idea"}
For most life histories, a trait acting on **juvenile survival** drives the fastest per-year
adaptation, with **fertility** close behind — because both feed directly into recruitment, the part
of the life cycle selection can turn over quickly. Selection on **adult survival** is the slowest to
register, especially in short-lived species. Hold this thought: in @sec-slow-fast it gets overturned.
:::

::: {.callout-warning title="Heads-up — `Beta` is on the link scale"}
The selection strength `Beta` lives on the logit/log scale of each vital rate, so a fixed `Beta = 0.15`
is **not** the same biological pressure across rates or species. That is deliberate — it is why the
original study reports several *definitions* of adaptation rate (@sec-under-the-hood) — but it means
you should compare scenarios *qualitatively* here, not read the bar heights as absolute selection.
:::

## Your turn — change the species

::: {.callout-important title="Exercise"}
Re-run the four scenarios for the **albatross** (species 5) by calling `run_scenario(rate, sp = 5)`
across `rates`. Does the ranking of vital rates change? Which rate now adapts fastest per year?
:::

::: {.callout-caution title="Solution" collapse="true"}
```{r}
#| label: solution-ch3
#| fig-height: 3.2
runs5 <- set_names(rates) |> map(\(r) run_scenario(r, sp = 5))

tibble(
  rate_label = vital_lab[rates],
  per_year   = map_dbl(runs5, "V")
) |>
  ggplot(aes(reorder(rate_label, per_year), per_year)) +
  geom_col(width = 0.65, fill = species_pal["5"]) +
  coord_flip() +
  labs(x = NULL, y = "Adaptation rate per year",
       title = "Albatross (species 5): the ranking shifts")
```

For the long-lived albatross, selection on **adult survival** is no longer last — in long-lived
species, adult survival carries much of the reproductive value, so a trait there matters more. The
"best" vital rate to target is **not** universal; it depends on where the species sits on the
slow–fast continuum, which is exactly the next chapter.
:::
