# The slow–fast continuum {#sec-slow-fast}

```{r}
#| include: false
source("R/_setup.R")
```

::: {.callout-note title="What you'll be able to do"}
Run the model across five species spanning the slow–fast life-history continuum, and show why
"which species adapts fastest?" has **no single answer** — it flips depending on whether you measure
adaptation per year or per generation.
:::

## Five species, one life cycle

The five species differ only in the four numbers that fill the Leslie matrix. Sorted by generation
time, they range from fast (a grass) to slow (an albatross):

```{r}
#| label: tbl-species
species |>
  transmute(Species = species, `Life history` = label,
            `Fertility F` = F, `Adult surv. SA` = SA, `Juv. surv. SJ` = SJ,
            `Maturation γ` = gamma, `Generation time T` = Tgen)
```

## Project every species under every selection scenario

We sweep all $5 \times 4 = 20$ combinations. The pattern is the same helper as before, mapped over a grid:

```{r}
#| label: full-sweep
project_one <- function(sp, rate) {
  MPM(
    h2 = 0.2, Vp = 1, Beta = 0.15,
    Fe = species$F[sp], SA = species$SA[sp], SJ = species$SJ[sp],
    S0 = species$SJ[sp], Y = species$gamma[sp],
    selectedrate = rate
  )
}

res <- expand_grid(sp = 1:5, rate = c("Fe", "SJ", "SA", "Y")) |>
  mutate(
    run      = map2(sp, rate, project_one),
    per_year = map_dbl(run, "V"),          # RAT: adaptation per year
    Tgen     = map_dbl(run, "T_GEN"),
    per_gen  = per_year * Tgen             # RAG: adaptation per generation
  ) |>
  select(-run) |>
  mutate(sp = factor(sp))
```

## The same result, two definitions

Now the heart of the matter. We plot the realised adaptation rate two ways — **per year** and **per
generation** — and the story changes completely.

```{r}
#| label: fig-two-definitions
#| fig-cap: "Adaptation rate across the slow–fast continuum (species 1 = fastest, 5 = slowest), under selection on each vital rate. Left: per year. Right: per generation. The same 20 runs tell two different stories."
#| fig-width: 8.5
#| fig-height: 4
res |>
  select(sp, rate, `per year` = per_year, `per generation` = per_gen) |>
  pivot_longer(c(`per year`, `per generation`),
               names_to = "unit", values_to = "rate_value") |>
  mutate(unit = factor(unit, levels = c("per year", "per generation"))) |>
  ggplot(aes(sp, rate_value, fill = rate)) +
  geom_col(position = position_dodge2(), width = 0.8) +
  facet_wrap(~unit, scales = "free_y") +
  scale_fill_manual(values = vital_pal, labels = vital_lab, name = "Selection on") +
  labs(x = "Species (fast → slow)", y = "Adaptation rate",
       title = "Who adapts fastest? It depends how you ask")
```

Read the two panels against each other:

- **Per year (left):** adaptation often slows from fast to slow species when selection acts on
  fecundity or juvenile survival. This pattern is not universal, however: when selection acts on
  maturation or adult survival, intermediate life histories can show the highest adaptation rates.
- **Per generation (right):** the ranking changes again. Adaptation decreases along the slow–fast
  continuum when selection acts on juvenile survival, remains constant across species when selection
  acts on fecundity, and increases with generation time when selection acts on maturation or adult
  survival.

## The cleanest demonstration: fecundity

Zoom in on selection acting on **fecundity** alone:

```{r}
#| label: fig-fertility-zoom
#| fig-cap: "Selection on fecundity only. Adaptation per year declines with generation time because the same evolutionary change is distributed across increasingly long generations. Theory predicts identical adaptation rates per generation across species; any small departures from a perfectly flat relationship reflect numerical approximation rather than biological differences (see @sec-under-the-hood)."
#| fig-height: 3.4
res |>
  filter(rate == "Fe") |>
  select(sp, `per year` = per_year, `per generation` = per_gen) |>
  pivot_longer(-sp, names_to = "unit", values_to = "rate_value") |>
  ggplot(aes(sp, rate_value, group = unit, colour = unit)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2.4) +
  scale_colour_manual(values = c("per year" = "#B5651D", "per generation" = "#3E6E83"),
                      name = NULL) +
  labs(x = "Species (fast → slow)", y = "Adaptation rate",
       title = "Adaptation under fecundity selection")
```

This pattern is predicted by the theory, not just by the simulation. When selection acts on
fecundity, the adaptation rate per generation reduces to $h^2 V_P\,\beta$, which contains no
life-history term. Theoretically, all species should therefore adapt at the same rate per generation.
Small deviations visible in the numerical results likely reflect discretisation or numerical
approximation. Per year, adaptation then declines with generation time because the same evolutionary
change is spread across more calendar years. We derive this result and compare it with the numerical
model in @sec-under-the-hood.

::: {.callout-tip title="Key idea — the punchline"}
There is **no universal answer** to "which species adapts fastest." It depends on (1) the species'
life cycle, (2) which vital rate selection targets, and (3) **how you define adaptation** — per year
or per generation, and in what units of selection. The original study makes exactly this point: don't
assume an answer from intuition; model the rate with the definition that matters to your question, and
read off the number [@vandewalle2025].
:::

::: {.callout-important title="Your turn"}
Your prediction from @sec-big-picture — grass versus albatross — was about adaptation *per year*.
Look at the left panel: were you right? Now look at the right panel. Did the albatross "lose" on both
definitions, or only one? Which definition would a conservation manager planning over **calendar
years** care about, and which would a quantitative geneticist reach for?
:::

::: {.callout-caution title="Discussion — there isn't one right answer" collapse="true"}
A manager allocating effort over the next few decades thinks in **years**, so the per-year panel is
their world: fast species adapt more in the time available. A quantitative geneticist studying the
*genetic* response thinks in **generations**, where the breeder's equation lives — and there the slow
albatross is far from hopeless. Both are correct; they are answering different questions.
:::

You have now done the core of the workshop: you can run the model, read it, and reason about its
central result. The next part opens the engine for those who want it.
