# The slow–fast continuum {#sec-slow-fast}

```{r}
#| include: false
source("R/_setup.R")
```

::: {.callout-note title="What you'll be able to do"}
Run the model across five species spanning the slow–fast life-history continuum, and show why
"which species adapts fastest?" has **no single answer** — it flips depending on whether you measure
adaptation per year or per generation.
:::

## Five species, one life cycle

The five species differ only in the four numbers that fill the Leslie matrix. Sorted by generation
time, they march from fast (a grass) to slow (an albatross):

```{r}
#| label: tbl-species
species |>
  transmute(Species = species, `Life history` = label,
            `Fertility F` = F, `Adult surv. SA` = SA, `Juv. surv. SJ` = SJ,
            `Maturation γ` = gamma, `Generation time T` = Tgen)
```

## Project every species under every selection scenario

We sweep all $5 \times 4 = 20$ combinations. The pattern is the same helper as before, mapped over a grid:

```{r}
#| label: full-sweep
project_one <- function(sp, rate) {
  MPM(
    h2 = 0.2, Vp = 1, Beta = 0.15,
    Fe = species$F[sp], SA = species$SA[sp], SJ = species$SJ[sp],
    S0 = species$SJ[sp], Y = species$gamma[sp],
    selectedrate = rate
  )
}

res <- expand_grid(sp = 1:5, rate = c("Fe", "SJ", "SA", "Y")) |>
  mutate(
    run      = map2(sp, rate, project_one),
    per_year = map_dbl(run, "V"),          # RAT: adaptation per year
    Tgen     = map_dbl(run, "T_GEN"),
    per_gen  = per_year * Tgen             # RAG: adaptation per generation
  ) |>
  select(-run) |>
  mutate(sp = factor(sp))
```

## The same result, two definitions

Now the heart of the matter. We plot the realised adaptation rate two ways — **per year** and **per
generation** — and the story changes completely.

```{r}
#| label: fig-two-definitions
#| fig-cap: "Adaptation rate across the slow–fast continuum (species 1 = fastest, 5 = slowest), under selection on each vital rate. Left: per year. Right: per generation. The same 20 runs tell two different stories."
#| fig-width: 8.5
#| fig-height: 4
res |>
  select(sp, rate, `per year` = per_year, `per generation` = per_gen) |>
  pivot_longer(c(`per year`, `per generation`),
               names_to = "unit", values_to = "rate_value") |>
  mutate(unit = factor(unit, levels = c("per year", "per generation"))) |>
  ggplot(aes(sp, rate_value, fill = rate)) +
  geom_col(position = position_dodge2(), width = 0.8) +
  facet_wrap(~unit, scales = "free_y") +
  scale_fill_manual(values = vital_pal, labels = vital_lab, name = "Selection on") +
  labs(x = "Species (fast → slow)", y = "Adaptation rate",
       title = "Who adapts fastest? It depends how you ask")
```

Read the two panels against each other:

- **Per year (left):** adaptation generally *slows down* as you move from fast to slow species — an
  albatross changes less per calendar year than a grass. But note the exception for adult survival,
  where intermediate species can peak.
- **Per generation (right):** the picture rearranges. For some vital rates the slow species now look
  *as adaptable or more so*, because each of their generations is worth many years.

## The cleanest demonstration: fertility

Zoom in on selection acting on **fertility** alone:

```{r}
#| label: fig-fertility-zoom
#| fig-cap: "Selection on fertility only. Per year, adaptation falls steeply with generation time. Per generation, it is essentially flat — every species adapts the same amount per generation."
#| fig-height: 3.4
res |>
  filter(rate == "Fe") |>
  select(sp, `per year` = per_year, `per generation` = per_gen) |>
  pivot_longer(-sp, names_to = "unit", values_to = "rate_value") |>
  ggplot(aes(sp, rate_value, group = unit, colour = unit)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2.4) +
  scale_colour_manual(values = c("per year" = "#B5651D", "per generation" = "#3E6E83"),
                      name = NULL) +
  labs(x = "Species (fast → slow)", y = "Adaptation rate",
       title = "Fertility selection: a textbook result")
```

This is not a quirk of the simulation — it falls straight out of the theory. When selection acts on
fertility, the per-generation rate reduces to $h^2 V_P\,\beta$, which contains **no species-specific
term at all**. Per year you then divide by generation time, so the slow species lag. (We derive and
check this in @sec-under-the-hood.)

::: {.callout-tip title="Key idea — the punchline"}
There is **no universal answer** to "which species adapts fastest." It depends on (1) the species'
life cycle, (2) which vital rate selection targets, and (3) **how you define adaptation** — per year
or per generation, and in what units of selection. The original study makes exactly this point: don't
assume an answer from intuition; model the rate with the definition that matters to your question, and
read off the number [@vandewalle2025].
:::

::: {.callout-important title="Your turn"}
Your prediction from @sec-big-picture — grass versus albatross — was about adaptation *per year*.
Look at the left panel: were you right? Now look at the right panel. Did the albatross "lose" on both
definitions, or only one? Which definition would a conservation manager planning over **calendar
years** care about, and which would a quantitative geneticist reach for?
:::

::: {.callout-caution title="Discussion — there isn't one right answer" collapse="true"}
A manager allocating effort over the next few decades thinks in **years**, so the per-year panel is
their world: fast species adapt more in the time available. A quantitative geneticist studying the
*genetic* response thinks in **generations**, where the breeder's equation lives — and there the slow
albatross is far from hopeless. Both are correct; they are answering different questions. Recognising
which definition a claim rests on is half the battle in cross-disciplinary work on adaptation.
:::

You have now done the core of the workshop: you can run the model, read it, and reason about its
central result. The next part opens the engine for those who want it.
