# Under the hood {#sec-under-the-hood}

```{r}
#| include: false
source("R/_setup.R")
source("R/theoretical_rates.R")
```

::: {.callout-note title="What you'll be able to do"}
Describe how the hyperstate matrix is assembled from small building blocks, see where inheritance
enters, list the six definitions of adaptation rate, and reproduce the paper's check that the
matrix model agrees with the analytic theory.
:::

This part is for the curious. None of it is needed to *use* `MPM()`, but understanding how the model
is constructed helps interpret its predictions, evaluate its assumptions, and apply it to new
biological questions.

## Three dimensions instead of one

A classic MPM tracks individuals by **stage** only — a vector of (juveniles, adults). The
EvoDemo-Hyper MPM tracks them by **stage × breeding value × phenotype**: a three-dimensional
*hyperstate* [@roth2016]. Each cell counts the individuals that are, say, "adult, with breeding value
class 12, expressing phenotype class 15".

The full projection matrix factorises into a survival part $\tilde{\mathbf{U}}$ and a reproduction
part $\tilde{\mathbf{F}}$,

$$
\mathbf{n}_{t+1} = \tilde{\mathbf{A}}[\mathbf{n}_t]\,\mathbf{n}_t,
\qquad
\tilde{\mathbf{A}} = \tilde{\mathbf{U}} + \tilde{\mathbf{F}},
$$

and each of those is built by applying one process per dimension, **shuffling the population vector
between steps with vec-permutation matrices** so that each block acts on the right axis
[@hunter2005]. The package handles all of this; the result is just sparse-matrix multiplication.

## The building blocks

The hyperstate matrix can be decomposed into a series of biologically distinct processes — survival,
reproduction, inheritance, and phenotype assignment. The conceptual organisation is what makes it
tractable; the underlying mathematics (block matrices, permutations, genetic transmission) is
genuinely sophisticated. Below are the small building blocks the model uses to implement these
processes, defined here so this chapter is self-contained:

```{r}
#| label: building-blocks-def
Emat <- function(i, j, u) {                 # a matrix of zeros with a single 1 at (i, j)
  E <- matrix(0, nrow = u, ncol = u)
  E[i, j] <- 1
  E
}

speye <- function(N) {                      # a sparse identity matrix
  Matrix::sparseMatrix(i = 1:N, j = 1:N, x = rep(1, N))
}

GVLE <- function(zz, Va) {                  # the Mendelian segregation kernel
  exp(-(zz^2) / Va) / sqrt(pi * Va)
}

BD_proj_mat <- function(B) {                # assemble a block-diagonal matrix from an array of blocks
  siz <- dim(B)[-1]
  s   <- prod(siz)
  sk  <- s / siz[1]
  siz <- siz[2:3]
  A   <- matrix(0, nrow = s, ncol = s)
  for (i in 1:sk) {
    r <- ((i - 1) %% siz[1]) + 1
    c <- floor((i - 1) / siz[1]) + 1
    A <- A + kronecker(Emat(i, i, sk), B[, , r, c])
  }
  A
}
```

`Emat()` places a single 1 in an otherwise empty matrix — the atom of the block-diagonal construction:

```{r}
#| label: emat
Emat(2, 3, 4)
```

`speye()` makes a sparse identity (breeding value and phenotype are assumed fixed within a lifetime,
so their within-life transition matrices are identities):

```{r}
#| label: speye
as.matrix(speye(3))
```

And `BD_proj_mat()` assembles a block-diagonal matrix from an array of blocks — the operation behind
$\mathbb{U}, \mathbb{B}, \mathbb{P}$ and friends:

```{r}
#| label: fig-block-diagonal
#| fig-cap: "A block-diagonal matrix built by `BD_proj_mat()` from a 2×2×3×3 array of blocks. This sparse structure is what makes the model fast."
#| fig-height: 3.6
toy <- array(runif(2 * 2 * 3 * 3, 1, 8), dim = c(2, 2, 3, 3))
image(BD_proj_mat(toy), main = "Block-diagonal structure")
```

## Where inheritance enters

Offspring do not inherit a parent's breeding value directly. Instead, their breeding value is drawn
from a distribution centred on the average breeding value of the two parents, with variance equal to
half the additive genetic variance — the **Fisher infinitesimal model**. The function `GVLE()`
defined above is exactly this segregation kernel:

```{r}
#| label: fig-segregation
#| fig-cap: "The Mendelian segregation kernel `GVLE()`: the spread of offspring breeding values around the mid-parent value, for Va = 0.2."
zz <- seq(-2, 2, by = 0.05)
tibble(z = zz, density = GVLE(zz, Va = 0.2)) |>
  ggplot(aes(z, density)) +
  geom_area(fill = species_pal["3"], alpha = 0.5) +
  geom_line(linewidth = 0.8, colour = species_pal["4"]) +
  labs(x = "Offspring breeding value (relative to mid-parent)", y = "Density",
       title = "Inheritance as a convolution")
```

Each year the model convolves the parental breeding-value distribution with this kernel — that single
step is the entire evolutionary engine.

## Six ways to define "adaptation rate"

A recurring theme of the workshop is that adaptation rate is not one number. The paper lays out
**six definitions** built from two choices:

- **Time unit** — per **year** (RAT) or per **generation** (RAG).
- **Selection unit** — on the linear/link scale ($\beta$), per unit of the **vital rate** ($\theta$),
  or per unit of **individual fitness** ($R_0$): giving $RA_{S\theta}$ and $RA_{SR_0}$ variants.

Their punchline: under the most standardised definition (per generation, per unit of fitness), the
rate collapses to $h^2 V_P$ — **identical for every species and every vital rate**. Strip away the
standardisation and the differences between species reappear. The "disagreement" between fields was
largely a disagreement about *units* [@vandewalle2025].

## Does the matrix model match the theory?

This is the validation that earns trust. We compute the analytic per-year rate (RAT) for all five
species under each vital rate, then overlay the *realised* rate from running `MPM()`.

```{r}
#| label: theory-table
theory <- species |>
  mutate(th = pmap(list(F, SA, SJ, Y = gamma),
                   \(F, SA, SJ, Y) theoretical_rates(F, SA, SJ, Y, Va = 0.2, beta = 0.15))) |>
  select(species, th) |>
  unnest(th) |>
  select(species, rate, theory_year = per_year)
```

```{r}
#| label: mpm-sweep-ch5
project_one <- function(sp, rate) {
  MPM(h2 = 0.2, Vp = 1, Beta = 0.15,
      Fe = species$F[sp], SA = species$SA[sp], SJ = species$SJ[sp],
      S0 = species$SJ[sp], Y = species$gamma[sp], selectedrate = rate)
}

mpm <- expand_grid(species = factor(1:5), rate = c("Fe", "SJ", "SA", "Y")) |>
  mutate(mpm_year = map2_dbl(as.integer(species), rate, \(s, r) project_one(s, r)$V))

compare <- left_join(mpm, theory, by = c("species", "rate"))
```

```{r}
#| label: fig-theory-vs-mpm
#| fig-cap: "Per-year adaptation rate: bars are the realised values from `MPM()`, points are the analytic theory. They line up across species and vital rates — the matrix model is doing what the maths says it should."
#| fig-width: 8.5
#| fig-height: 4
ggplot(compare, aes(species)) +
  geom_col(aes(y = mpm_year, fill = rate), alpha = 0.55, width = 0.7) +
  geom_point(aes(y = theory_year), size = 2, colour = "grey15") +
  facet_wrap(~rate, scales = "free_y", labeller = as_labeller(vital_lab)) +
  scale_fill_manual(values = vital_pal, guide = "none") +
  labs(x = "Species (fast → slow)", y = "Adaptation rate per year",
       title = "Simulation (bars) vs analytic theory (points)")
```

::: {.callout-tip title="Key idea"}
The matrix model is not a black box that *replaces* theory — it *agrees* with it, and then keeps
working in the many cases where closed-form formulas run out (non-Gaussian trait distributions,
evolving variances, plasticity). That is the whole argument for the approach.
:::

## Tuning and the artefacts to watch

A few arguments control the numerical grid. Used carelessly they produce believable-looking nonsense,
so it pays to know them.

- **`g`** — number of breeding-value/phenotype classes (default 20). More classes = finer resolution
  but slower (≈ 1 s at `g = 10`, ≈ 8 s at `g = 40`). The paper gives a guideline for the minimum
  needed for accurate discretisation, $b \ge 4/\sqrt{h^2 V_P} \approx 9$ at the default genetics, and
  used 40 classes.
- **`xp`** — the half-range of trait values, i.e. the grid spans $[-x_p, x_p]$ (default 4).
- **`timeA`** — number of years projected (default 100).

::: {.callout-warning title="Heads-up — the boundary artefact"}
If the trait adapts far enough to reach the edge of the grid, the distribution piles up against the
boundary and the adaptation rate becomes unreliable. The fix is to give it more room (`xp`) or fewer
years (`timeA`). The model **warns you** when the genetic distribution is too narrow or too wide for
the chosen grid — read those warnings.
:::

```{r}
#| label: boundary-warning
#| warning: true
#| fig-height: 3
# A deliberately under-resolved setup: very low heritability on a coarse grid.
risky <- MPM(g = 20, xp = 2, h2 = 0.01, Beta = 0.15,
             Fe = species$F[4], SA = species$SA[4], SJ = species$SJ[4],
             S0 = species$SJ[4], Y = species$gamma[4],
             enforce_initial_lambda1 = TRUE)

plot(risky$properties_per_year[, c("year", "Va")], type = "l",
     xlab = "year", ylab = "realised Va",
     main = "Additive variance eroding — partly real, partly artefact")
```

The realised additive variance drifts away from the expected value: some of this is genuine erosion
of variance under selection, but some is the grid being too coarse to represent a narrow
distribution. Always check `Va` and `Vp` in `properties_per_year` against what you asked for.

Two more useful switches:

- **`enforce_initial_lambda1 = TRUE`** adjusts fertility so the population starts at $\lambda = 1$
  (a stationary baseline) — handy when you want to isolate evolution from raw growth.
- **`stabilise_population_structure = FALSE`** starts away from the stable stage distribution, letting
  you observe transient demographic oscillations before equilibrium.
