# Before you arrive {.unnumbered}

Please run this once **before** the workshop. It takes ~10 minutes and means we start by modelling,
not troubleshooting.

## 1. R and a few CRAN packages

You need R (≥ 4.2) and, ideally, RStudio or Positron. Install the helper packages:

```{r}
#| eval: false
install.packages(c("remotes", "tidyverse", "pracma", "gsignal", "purrr", "tidyr", "Matrix"))
```

## 2. The workshop package

`evodemohypermpm` lives on the CNRS *Koda* forge, not CRAN, so we install it straight from there:

```{r}
#| eval: false
remotes::install_gitlab(
  repo = "timotheebonnet/evodemohypermpm",
  host = "https://src.koda.cnrs.fr/"
)
```

::: {.callout-note title="If install_gitlab cannot reach Koda"}
Some institutional networks block the forge. As a fallback, clone the repository and install locally:

```{r}
#| eval: false
# in a terminal:  git clone https://src.koda.cnrs.fr/timotheebonnet/evodemohypermpm.git
remotes::install_local("evodemohypermpm")
```
:::

## 3. Smoke test

Load the package and run one tiny projection. If you get a plot, you are ready.

```{r}
#| eval: false
library(evodemohypermpm)

test <- MPM(g = 10, Fe = 0.2286, SA = 0.95, SJ = 0.8, S0 = 0.8, Y = 0.07)
plot(test$properties_per_year[, c("year", "N")], type = "l",
     xlab = "year", ylab = "population size")
```

::: {.callout-note title="One detail to check"}
After installation, the package should load as `library(evodemohypermpm)` and expose the function
`MPM()`. If `library()` complains, open the project's `DESCRIPTION` file and use whatever name appears
on the `Package:` line. Everything else in this workshop is unchanged.
:::

## Reproducibility

This whole site is built from plain `.qmd` files with [Quarto](https://quarto.org). Two things make
your results reproducible:

- **`R/_setup.R`** loads the packages, sets the shared plotting theme, and defines the five-species
  table. Every chapter sources it, so a colour or a parameter always means the same thing.
- We record the exact computing environment below. If you ever get different numbers, compare your
  session to this one first.

If you want to *lock* the package versions, the repository ships with an optional
[`renv`](https://rstudio.github.io/renv/) lockfile — run `renv::restore()` to reproduce the exact
environment used to build the site.

```{r}
#| eval: false
sessionInfo()
```
